---
title: "SRS: A Subject Randomization System"
author: "Balasubramanian Narasimhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: srs.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction


SRS is a _Subject Randomization System_ based on the paper by Pocock
and Simon [-@ps:1975]. It follows the development
in the paper rather closely. In this vignette we show how one might
use the system in designing and implementing randomizations for
clinical trials.

__Note__: Although everything here should work, this is somewhat dated
code that I put on R-forge and forgot for a good bit of time. I'd do
this much differently now; in fact, I have a version using `R6`
classes that I hope to release at some point.

This vignette has two parts to it. The first part goes into detail
discussing some of the innards of the package. This is most meaningful
to those in our Biostatistics core who may recommend this software for
use in trials.  The second part is more information on how one might
use it in conducting a trial.

This package is written using `S4` classes. No deep knowledge of S4
classes is assumed in what follows.

---

## The Basic Classes

There are two main classes that most users of the package will use:
`ClinicalExperiment` and `PocockSimonRandomizer`. The class
`ClinicalExperiment`, as the name implies, encapsulates the
characteristics of a clinical experiment. An instance of this class is
used to create an instance of the other class `PocockSimonRandomizer`
so that the randomizer remains associated with a particular clinical
experiment.

---

## A Simple Example

### A Clinical Experiment

Let us create a simple clinical experiment object after invoking the
requisite package. The function `ClinicalExperiment` (as distinct from
the `ClinicalExperiment` class) is available for us.

```{r}
library(SRS)
expt0 <- ClinicalExperiment(number.of.factors = 3,
                            number.of.factor.levels = c(2, 2, 3),
                            number.of.treatments = 3)
```

This create an experiment with three factors and three treatments. The
first factor has 2 levels, the second 2, and the third 3. If none of
the arguments are specified, the default is to create a two-factor,
two-treatment experiment with each factor having two levels. One can
name the factors with the argument `factor.names` but default names
such as `F_1`, `F_2`, etc., are provided. The levels are currently
indicated by the suffixes `-1`, `-2`, etc., that are attached to the
factor names; a flexible naming scheme for this might be introduced
later.

It is useful to print the object to see what it contains.
```{r}
print(expt0)
```

Of course, in anything other than a toy setting, one actually provides
some names for the factor and factor levels. We'll use this in what
follows.

```{r}
expt <- ClinicalExperiment(number.of.factors = 3,
                           factor.names = c("Sex", "Race", "Stage"),
                           number.of.factor.levels = c(2, 2, 3),
                           factor.level.names =
                               list(c("Female", "Male"),
                                    c("Caucasian", "Non-caucasian"),
                                    c("I", "II", "III")),
                           number.of.treatments = 3,
                           treatment.names <- c("Placebo", "Arm1", "Arm2"))
print(expt)
```

### The PocockSimon Randomizer

Now let's create a randomizer that will work for this experiment.

```{r}
r.obj <- new("PocockSimonRandomizer", expt, 12345L)
print(r.obj)
```

Note that we don't have a helper constructor function (for no
particular reason) and so we had to use the `new` function to create
the object. (Indeed, that is what the `ClinicalExperiment` function
does behind the scenes.)

The output of the print above indicates that there are some default
settings for the randomizer. For example, the treatment ratios are all
`1`'s indicating equal treatment preference; others such as `1 2 1`
could have been specified. Note the `stateTable` slot which will
summarize the margins of the factor distributions by treatment. Since
no randomization has been done, the slot `tr.assignments` is empty.

Of interest are the slots named `d.func`, `g.func` and `p.func`. The
`d.func` computes imbalance due to assigning each of the treatments,
`g.func` computes the overall imbalance, and the `p.func` computes the
probabilities of assigning each treatment based on the overall
imbalance. All of these can be changed by the user. Default values for
these functions are the ones described in [@ps:1975].

\subsection{Using the Randomizer}
\label{sec:using}

Now that we have defined the experiment and the randomizer, we can
randomize several subjects using these classes. First some helper
functions that are useful in simulations for generating random
identifiers and factors.

```{r}
###
### Generate a random Id for a subject (max 10000000)!
###
generateId <- function(i) {
  if (i < 0 || i > 10000) {
    stop("generateId: Arg expected to be between 1 and 9999")
  }
  zero.count <- 5 - trunc(log10(i)) - 1
  prefix <- substring(10^zero.count, 2)
  paste("ID.", prefix, i, sep="")
}

###
### Generate random factors; if n is the number of factors, limits is a list
### of length n with each element being a vector of possible factor levels
###
generateRandomFactors <- function(factor.levels) {
  unlist(lapply(factor.levels, function(x) sample(x, 1)))
}
```

Now, we will run a 10 randomizations and print the results.

```{r}
for (i in seq(10))
  r.obj <- randomize(r.obj, generateId(i),
                     generateRandomFactors(expt@factor.level.names))
print(r.obj@tr.assignments)
```

Just in case we are only interested in the last assigned treatment:

```{r}
lastRandomization(r.obj)
```

We can also look at the marginal distributions thus:

```{r}
print(r.obj@stateTable)
```

---

## Customizing the Randomizer

The functions for computing imbalance, overall imbalance and
probabilities can all be customized. These are best illustrated by
additional examples.

### A different imbalance function

Let's move away from the default range function to say the standard
deviation (`sd`) function.

```{r}
r.obj.2 <- new("PocockSimonRandomizer", expt, as.integer(12345),
               d.func = sd)
print(r.obj.2@d.func)
```

Now let's run that simulation again.

```{r}
for (i in seq(10))
    r.obj.2 <- randomize(r.obj.2, generateId(i),
                         generateRandomFactors(expt@factor.level.names))
print(r.obj.2@tr.assignments)
```

Now print the summaries.

```{r}
print(table(r.obj@tr.assignments[, "Treatment"]))
print(r.obj@stateTable)
```

### Weighting factors differently

Now let's weight imbalance on factor 1 more than the others by a
factor of 5. We do this by modifying the `g.func`.

```{r}
## Note: imbalances is a number of factors by number of treatments matrix
g.func <- function(imbalances) {
    factor.weights <- c (5, 1, 1)
    imbalances %*% factor.weights
}

r.obj.3 <- new("PocockSimonRandomizer", expt, 12345L,
               d.func = sd, g.func = g.func)
print(r.obj.3@g.func)
```

Now the simulation.

```{r}
for (i in seq(1000))
    r.obj.3 <- randomize(r.obj.3, generateId(i),
                         generateRandomFactors(expt@factor.level.names))
```

Let's look at the distribution of treatments and the marginal
distribution of factors.

```{r}
print(table(r.obj.3@tr.assignments[, "Treatment"]))
print(r.obj.3@stateTable)
```

### Unequal treatment assignments

Next, we try a simulation where we require 5:2:1 randomization. To
really see the effect, we need to change the function that computes
probabilities for picking each treatment based on the
randomization. Let's be greedy and use the following:

```{r}
p.func.greedy <- function(overallImbalance) {
    number.of.treatments <- length(overallImbalance)
    k <- which(overallImbalance == min(overallImbalance))
    ## Note there could be ties here...
    p.vec <- rep(0, number.of.treatments)
    p.vec[k] <- 1
    p.vec / sum(p.vec) ## will pick ties randomly
}
```

Now, a new randomizer.

```{r}
r.obj.4 <- new("PocockSimonRandomizer", expt, 12345L,
               tr.ratios = c(5, 2, 1), p.func = p.func.greedy)
```

A simulation.

```{r}
for (i in seq(1000))
    r.obj.4 <- randomize(r.obj.4, generateId(i),
                         generateRandomFactors(expt@factor.level.names))
```

```{r}
print(table(r.obj.4@tr.assignments[, "Treatment"]))
print(r.obj.4@stateTable)
```

### A different probability assignment

The drawback of using the greedy function in the previous example is
that there is some predictability as to what the randomizer will
assign based on the current state.  To throw in a bit of uncertainty,
we can define another function that favors the appropriate treatment
heavily, but not deterministically.

```{r}
p.func.not.so.greedy <- function(overallImbalance) {
    FAVORED.PROB <- 0.75
    number.of.treatments <- length(overallImbalance)
    k <- which(overallImbalance == min(overallImbalance))
    if (length(k) > 1) {
        k <- sample(k, 1)
    }
    p.vec <- rep((1 - FAVORED.PROB) / (number.of.treatments - 1), number.of.treatments)
    p.vec[k] <- FAVORED.PROB
    p.vec
}
```

```{r}
r.obj.5 <- new("PocockSimonRandomizer", expt, 28923L,
               tr.ratios = c(5, 2, 1), p.func = p.func.not.so.greedy)
```

Now, a simulation.

```{r}
for (i in seq(1000))
    r.obj.5 <- randomize(r.obj.5, generateId(i),
                         generateRandomFactors(expt@factor.level.names))
```

```{r}
print(table(r.obj.5@tr.assignments[, "Treatment"]))
print(r.obj.5@stateTable)
```

Another possibility for the probability function might be based on the
actual imbalances.

```{r}
p.func.imbalance <- function(overallImbalance) {
    p.vec <- overallImbalance / sum(overallImbalance)
    p.vec
}
```

Of course, this assumes that the imbalances calculated are
non-negative, which would be the case with range or standard
deviation. But some care must be taken to ensure this is the case for
arbitrary situations.

---

## Notes

The current package can be used without recourse to a database for
persistence.  This would require the initial definition of the
clinical experiment as in the example(s) above along with the
randomizer. This is done once for a study on a designated computer
running R to which the person assigned to do the randomization will
have primary access.

Thereafter, every time a subject is to be randomized (after all the
usual procedures for registration in the study) the randomization
process will require merely an id for the subject and the levels of
the prognostic factors of interest. The randomization is performed
simply by running the code snippet

```{r, eval = FALSE}
r.obj <- randomize(r.obj, id, c(fac1, fac2, fac3))
lastRandomization(r.obj)
```

where `r.obj` is a randomizer created as above, and `id`, `fac1`,
`fac2`, `fac3`, are the study id and the associated factor levels of
the subject to be randomized.

After each assignment, the person can save the R workspace so that the
state is preserved.  If R is invoked from the same directory again,
the state is restored for subsequent randomizations. Of course, this
means all the usual responsibilities for saving the workspace apply
for this mode of operation.

---

### References
